% Davis-Kahan sin-Theta: what the Lean formalization proves, and how it relates
% to the literature.  Written as an expert-readable companion to
% ForMathlib/Analysis/InnerProductSpace/DavisKahan.lean, so that the statement
% can be checked against the source papers without reading Lean.
%
% Independently written; the two source-paper digests it draws on are
%   ForMathlib/prose/Davis-1963-core-arguments.tex
%   ForMathlib/prose/Yu-Wang-Samworth-2014-core-arguments.tex

\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{booktabs}
\usepackage[colorlinks=true, urlcolor=blue, citecolor=blue, linkcolor=black]{hyperref}

\newcommand{\norm}[1]{\left\lVert #1\right\rVert}
\newcommand{\eps}{\varepsilon}
\newcommand{\F}{\mathrm{F}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\inner}[2]{\langle #1,\,#2\rangle}
\newcommand{\gap}{g}
\newcommand{\HH}{\mathcal H}
\newcommand{\sinT}{\sin\Theta}
\newcommand{\tp}{^{\mathsf T}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\ran}{ran}
\DeclareMathOperator{\spann}{span}

\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\title{Davis--Kahan $\sin\Theta$: what the Lean formalization proves,\\
and how it relates to Davis (1963/1970) and Yu--Wang--Samworth (2015)}

% NOTE: if you are a different model making a substantial addition and are not
% yet credited, add yourself to the author list above.
\author{Jonathan Crall \and Claude Opus 4.8}

\date{}

\begin{document}
\maketitle

\begin{abstract}
This note states, in ordinary mathematical language, the Davis--Kahan (DK)
$\sin\Theta$ theorem formalized in Lean, and locates it among the source papers.
The formalization is a ladder of Frobenius-norm $\sin\Theta$ bounds built on one
estimate, whose sharpest rung is $\norm{\sinT}_{\F}\le\norm{S-T}_{\F}/\gap$ with a
dimension-free constant.  The estimate is Davis's elementary cross-term /
Rayleigh-quotient inequality, and the separation is a hybrid gap between the
leading eigenvalues of one operator and the trailing eigenvalues of the other.
The hybrid-gap bound is the deterministic core of the Yu--Wang--Samworth (YWS)
variant; recovering their population-only gap is a further Weyl step, also
formalized.  A closing section records the DK variants left unformalized and what
each of them requires.
\end{abstract}

\vspace{1.5em}
% NOTE: Update permalink if underlying proof changes, so this always is up to date wrt to the referenced lean file.
\noindent Companion note to \href{https://github.com/AIQ-Kitware/aiq-dkps-formalization/blob/375251df7ab712397977635e8f607c11c3a8083d/ForMathlib/Analysis/InnerProductSpace/DavisKahan.lean}{ForMathlib/Analysis/InnerProductSpace/DavisKahan.lean}, part of a larger formalization effort at \href{https://github.com/AIQ-Kitware/aiq-dkps-formalization}{github.com/AIQ-Kitware/aiq-dkps-formalization}.

\section{Setup and notation}

Let $\HH$ be a finite-dimensional inner product space over
$\mathbb K\in\{\mathbb R,\mathbb C\}$ with $\dim\HH=n$, and let $T,S:\HH\to\HH$ be
self-adjoint.  Write the eigenvalues and an orthonormal eigenbasis of $T$ as
$\lambda_1(T),\dots,\lambda_n(T)$ and $u_1,\dots,u_n$, and those of $S$ as
$\lambda_1(S),\dots,\lambda_n(S)$ and $v_1,\dots,v_n$.  Fix a block size $d$; in
the intended application indices $1,\dots,d$ are the top $d$ eigenvalues, while
the statements below use only the split $\{1,\dots,d\}$ against
$\{d+1,\dots,n\}$, so sorting is not required.

Let $P$, $\widehat P$, and $Q=I-\widehat P$ be the orthogonal projections onto,
respectively,
\[
  \spann(u_1,\dots,u_d),\qquad
  \spann(v_1,\dots,v_d),\qquad
  \spann(v_{d+1},\dots,v_n).
\]
The first two are the leading spectral subspaces of $T$ and $S$, and $\ran Q$ is
the trailing $S$-subspace.  Let $\Theta$ be the vector of principal angles between
$\ran P$ and $\ran\widehat P$.  Two norms appear: the operator norm
$\norm{\cdot}_{\op}$ and the Frobenius, or Hilbert--Schmidt, norm
$\norm{B}_{\F}^2=\tr(B^*B)=\sum_k\norm{Be_k}^2$ over any orthonormal basis $e_k$.

The separation hypothesis is a \emph{hybrid gap}: a number $\gap>0$ with
\begin{equation}\label{eq:gap}
  \gap\le\bigl|\lambda_i(T)-\lambda_j(S)\bigr|
  \qquad\text{for all }i\le d<j.
\end{equation}
It pairs the leading eigenvalues of $T$ with the trailing eigenvalues of $S$, the
separation that Davis's estimate consumes.

\section{The formalized theorem}

The overlap $\sum_{i\le d}\sum_{j>d}|\inner{u_i}{v_j}|^2$ equals
$\norm{\sinT}_{\F}^2$, and the bounds on it form one chain of inequalities.

\begin{theorem}[Frobenius $\sin\Theta$ ladder]\label{thm:ladder}
Let $T,S$ be self-adjoint and let $\gap>0$ satisfy \eqref{eq:gap}.  Then
\begin{equation}\label{eq:ladder}
  \gap^2\,\norm{\sinT}_{\F}^2
  \;\le\;\norm{P(S-T)Q}_{\F}^2
  \;\le\;\norm{(S-T)P}_{\F}^2
  \;\le\;\norm{S-T}_{\F}^2,
\end{equation}
and consequently
\[
  \norm{\sinT}_{\F}\le\frac{\norm{S-T}_{\F}}{\gap},
  \qquad
  \norm{\widehat P-P}_{\F}\le\frac{\sqrt2\,\norm{S-T}_{\F}}{\gap},
\]
with constants independent of $n$.
\end{theorem}

The three right-hand sides of \eqref{eq:ladder} are the leading$\times$trailing
block of the perturbation, its restriction to the leading subspace (the one-sided
residual of the original DK theorem), and its full Frobenius norm; dividing by
$\gap^2$ gives three overlap bounds, of which the last is sharp.  The projector
distance and the overlap differ by an exact factor,
$\norm{\widehat P-P}_{\F}^2=2\norm{\sinT}_{\F}^2$, because
$\widehat P-P=\widehat P(I-P)-(I-\widehat P)P$ splits into two pieces with
orthogonal ranges of equal Frobenius norm, each equal to the overlap.

\begin{corollary}[operator-norm form]\label{cor:opnorm}
If $\norm{S-T}_{\op}\le\eps$, the residual has $d$ nonzero columns of norm at most
$\eps$, so $\norm{(S-T)P}_{\F}^2\le d\,\eps^2$ and
$\norm{\sinT}_{\F}^2\le d\,\eps^2/\gap^2$.  Extending to all $n$ columns gives the
cruder $\norm{\sinT}_{\F}^2\le n\,\eps^2/\gap^2$ and
$\norm{\widehat P-P}_{\F}^2\le 2n\,\eps^2/\gap^2$.  The example $S-T=\eps I$ shows
the dimension factor is unavoidable once only $\norm{S-T}_{\op}$ is controlled.
\end{corollary}

\begin{corollary}[rank and eigenvalue floor: the statistical setup]\label{cor:floor}
If $T$ is positive semidefinite of rank $d$ with nonzero eigenvalues at least
$\alpha$, and $\norm{S-T}_{\op}\le\eps\le\alpha/2$, then Weyl's inequality places
every trailing eigenvalue of $S$ below $\alpha/2$, so \eqref{eq:gap} holds with
$\gap=\alpha/2$ and $\norm{\sinT}_{\F}^2\le 4n\,\eps^2/\alpha^2$.
\end{corollary}

\section{The proof mechanism}

For eigenvectors $u_i$ of $T$ and $v_j$ of $S$, self-adjointness gives the
\emph{cross-term identity}
\begin{equation}\label{eq:cross}
  \inner{u_i}{(S-T)v_j}=\bigl(\lambda_j(S)-\lambda_i(T)\bigr)\inner{u_i}{v_j}.
\end{equation}
On a cross pair ($i\le d<j$), \eqref{eq:gap} turns this into
$\gap^2\,|\inner{u_i}{v_j}|^2\le|\inner{u_i}{(S-T)v_j}|^2$.  Summing over the
cross block gives the first inequality of \eqref{eq:ladder}, with right-hand side
$\norm{P(S-T)Q}_{\F}^2$; enlarging the column sum and then the row sum, by
Parseval in the bases $\{v_j\}$ and $\{u_i\}$, produces the remaining two rungs
and the full $\norm{S-T}_{\F}^2$.  The argument is elementary throughout, with no
resolvents or contour integrals.

\section{Relation to Davis (1963/1970)}

The estimate is Davis's~\cite{Davis1963}.  For a simple spectrum his
total-rotation bound reads, with $A\rightsquigarrow T$, $A+H\rightsquigarrow S$,
and $\mathcal C^{\perp}U$ the off-diagonal part of the matching unitary,
\[
  (\gamma')^2\,\norm{\mathcal C^{\perp}U}_{\F}^2
  \le\norm{H}_{\F}^2-\sum_i\bigl(\lambda_i-\lambda_i'\bigr)^2,
  \qquad
  \gamma_i=\min_{j\ne i}|\lambda_i-\lambda_j'|,
\]
proved by comparing $\inner{(A+H-\lambda_i')^2x_i}{x_i}$ from two sides, which is
\eqref{eq:cross}.  The formalization runs this argument on the two-block split.
The ladder \eqref{eq:ladder} realizes Davis's off-diagonal principle: the
cross-block rung $\norm{P(S-T)Q}_{\F}^2$ and the residual rung
$\norm{(S-T)P}_{\F}^2$ control the rotation by the block-mixing part of $H$,
and the sharp rung relaxes this to $\norm{S-T}_{\F}^2$.  The hybrid gap
\eqref{eq:gap} is Davis's $\gamma_i$ restricted to the cross block.  The companion
part~III of Davis and Kahan~\cite{DavisKahan1970} states the $\sin\Theta$ theorem
for every unitarily invariant norm; the Frobenius case is the instance formalized
here.

\section{Relation to Yu--Wang--Samworth (2015)}

Yu, Wang and Samworth~\cite{YWS2015} work with symmetric $\Sigma,\widehat\Sigma$
and a population eigengap $\Delta=\min(\lambda_{r-1}-\lambda_r,\
\lambda_s-\lambda_{s+1})$ formed from $\Sigma$ alone, and prove
\[
  \norm{\sinT(\widehat V,V)}_{\F}
  \le\frac{2\min\{d^{1/2}\norm{\widehat\Sigma-\Sigma}_{\op},
                   \norm{\widehat\Sigma-\Sigma}_{\F}\}}{\Delta}.
\]
Their proof is a residual sandwich
$\Delta\norm{\sinT}_{\F}\le\norm{R}_{\F}\le
2\min\{d^{1/2}\norm{\cdot}_{\op},\norm{\cdot}_{\F}\}$ with
$R=\widehat V\Lambda-\Sigma\widehat V$: a deterministic separation estimate on the
left, a perturbation estimate (Weyl and Hoffman--Wielandt) on the right, and the
factor $2$ paid for moving to the population-only gap $\Delta$.

The formalization keeps these two pieces separate.  The hybrid-gap bound, the
sharp rung of \eqref{eq:ladder}, is the deterministic separation estimate with
constant $1$: the hybrid gap already places the trailing eigenvalues, so the
Hoffman--Wielandt displacement term drops out.  The lemma
\texttt{gap\_of\_eigengap} supplies the Weyl step, converting a spectral gap of
$T$ (leading eigenvalues $\ge a$, trailing $\le b$) into the hybrid gap
$(a-b)-\eps$; YWS's $\Delta=\lambda_d-\lambda_{d+1}$ is the case $a-b$, and
\texttt{gap\_of\_rank\_floor} the case $a=\alpha$, $b=0$.  Their $d^{1/2}$
operator-norm branch is Corollary~\ref{cor:opnorm},
$\norm{\sinT}_{\F}\le d^{1/2}\eps/\gap$.

\section{What is not formalized}

The formalization covers the Frobenius $\sin\Theta$ theorem with a hybrid or
population gap.  The remaining DK-family results, and the machinery each would
require, are as follows.

\begin{itemize}
\item \emph{Operator-norm and general unitarily-invariant-norm $\sin\Theta$.} The
  dimension-free $\norm{\sinT}_{\op}\le\norm{S-T}_{\op}/\gap$, and its extension
  to every unitarily invariant norm (the full part~III statement).  The
  sum-of-squares route delivers the Frobenius norm; the operator-norm version
  needs the largest singular value of the residual, and the general one needs the
  theory of unitarily invariant norms (symmetric gauge functions, Ky Fan
  majorization), which Mathlib does not yet carry.
\item \emph{Davis's sharper total-rotation estimate.} The refinement that
  subtracts the eigenvalue displacement $\sum_i(\lambda_i-\lambda_i')^2$, and its
  $\norm{\mathcal C^{\perp}H}_{\F}$ form.  These rest on the canonical matching
  unitary, built from the polar factor of $P_j\widehat P_jP_j$, together with
  Davis's Schur--Horn eigenvalue-change lower bound.
\item \emph{The $\tan\Theta$, $\sin2\Theta$, and $\tan2\Theta$ theorems.} The
  other members of the DK family, each with its own gap condition and $2\times2$
  compression argument.
\item \emph{The YWS aligned-basis bound.} The orthogonal Procrustes alignment: the
  orthogonal $O$ minimizing $\norm{\widehat V O-V}_{\F}$, for which
  $\norm{\widehat V O-V}_{\F}\le 2^{3/2}\min\{\cdots\}/\Delta$.  It needs the
  singular value decomposition of the cross-Gram $\widehat V\tp V$ and the
  identity $2d-2\sum_k\cos\theta_k\le 2\sum_k\sin^2\theta_k$.
\item \emph{The YWS singular-vector extension.} The bound for singular subspaces
  of rectangular $A,\widehat A$, obtained by applying the symmetric result to
  $A\tp A$; it needs singular value perturbation and a Gram-perturbation bound on
  $\norm{\widehat A\tp\widehat A-A\tp A}$.
\item \emph{General-interval subspaces.} Spectral subspaces cut out by an
  arbitrary interval, with a two-sided gap, rather than the sorted leading block
  used here.
\end{itemize}

\section{Dictionary to the Lean source}

All declarations live in
\texttt{ForMathlib/Analysis/InnerProductSpace/DavisKahan.lean}
(namespace \texttt{ForMathlib}).

\begin{center}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}p{0.42\linewidth}l@{}}
\toprule
Statement here & Lean name \\
\midrule
Cross-term identity \eqref{eq:cross}
  & \texttt{inner\_eigenvectorBasis\_map\_sub\_eigenvectorBasis} \\
Parseval identities (full sum / single row)
  & \texttt{sum\_sq\_norm\_inner\_\dots\_map\_sub\_eq}\,/\,\texttt{\dots\_eq\_row} \\
Dimension-factor step $\norm{S-T}_{\F}^2\le n\eps^2$
  & \texttt{sum\_norm\_eigenvectorBasis\_map\_sub\_sq\_le} \\
Ladder, cross-block rung $\norm{P(S-T)Q}_{\F}^2/\gap^2$
  & \texttt{sum\_cross\_\dots\_sq\_le\_offDiag} \\
Ladder, residual rung $\norm{(S-T)P}_{\F}^2/\gap^2$
  & \texttt{sum\_cross\_\dots\_sq\_le\_residual} \\
Ladder, sharp rung $\norm{S-T}_{\F}^2/\gap^2$ (Thm~\ref{thm:ladder})
  & \texttt{sum\_cross\_\dots\_sq\_le\_hilbertSchmidt} \\
Cor.~\ref{cor:opnorm}, $d^{1/2}$ branch $d\,\eps^2/\gap^2$
  & \texttt{sum\_cross\_\dots\_sq\_le\_opNorm} \\
Cor.~\ref{cor:opnorm}, crude $n\eps^2/\gap^2$
  & \texttt{sum\_cross\_\dots\_sq\_le} \\
Population-gap Weyl bridge $\Rightarrow\gap=(a-b)-\eps$
  & \texttt{gap\_of\_eigengap} \\
Projector identity $\norm{\widehat P-P}_{\F}^2=2\cdot\text{overlap}$
  & \texttt{sum\_norm\_sub\_starProjection\_span\_sq\_eq} \\
Projector form of Thm~\ref{thm:ladder}
  & \texttt{sum\_norm\_sub\_starProjection\_span\_sq\_le\_hilbertSchmidt} \\
Projector, crude $2n\eps^2/\gap^2$
  & \texttt{sum\_norm\_sub\_starProjection\_span\_sq\_le} \\
Weyl bridge, rank/floor $\Rightarrow\gap=\alpha/2$
  & \texttt{gap\_of\_rank\_floor} \\
Cor.~\ref{cor:floor}, rank/floor $4n\eps^2/\alpha^2$
  & \texttt{sum\_cross\_\dots\_sq\_le\_of\_rank\_floor} \\
\bottomrule
\end{tabular}
\end{center}

The projections $P,\widehat P$ are Mathlib's \texttt{Submodule.starProjection}
of the spans of the selected eigenvectors; the Frobenius norm on the left of the
projector bound is written in Lean as the basis sum
$\sum_k\norm{(\widehat P-P)u_k}^2=\norm{\widehat P-P}_{\F}^2$.

\begin{thebibliography}{9}
\bibitem{Davis1963}
Chandler Davis.
\newblock \href{https://doi.org/10.1016/0022-247X(63)90001-5}{The rotation of eigenvectors by a perturbation}.
\newblock \emph{J.\ Math.\ Anal.\ Appl.}\ \textbf{6} (1963), 159--173.

\bibitem{DavisKahan1970}
Chandler Davis and W.\ M.\ Kahan.
\newblock \href{https://doi.org/10.1137/0707001}{The rotation of eigenvectors by a perturbation.\ III}.
\newblock \emph{SIAM J.\ Numer.\ Anal.}\ \textbf{7} (1970), no.\ 1, 1--46.

\bibitem{YWS2015}
Yi Yu, Tengyao Wang, and Richard J.\ Samworth.
\newblock \href{https://doi.org/10.1093/biomet/asv008}{A useful variant of the Davis--Kahan theorem for statisticians}.
\newblock \emph{Biometrika} \textbf{102} (2015), no.\ 2, 315--323.
\newblock arXiv:\href{https://arxiv.org/abs/1405.0680}{1405.0680}.
\end{thebibliography}

\end{document}
