% Davis-Kahan sin-Theta: what the Lean formalization proves, and how it relates
% to the literature.  Written as an expert-readable companion to
% ForMathlib/Analysis/InnerProductSpace/DavisKahan.lean, so that the statement
% can be checked against the source papers without reading Lean.
%
% Independently written; the two source-paper digests it draws on are
%   ForMathlib/prose/Davis-1963-core-arguments.tex
%   ForMathlib/prose/Yu-Wang-Samworth-2014-core-arguments.tex

\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{booktabs}
\usepackage[colorlinks=true, urlcolor=blue, citecolor=blue, linkcolor=black]{hyperref}

\newcommand{\norm}[1]{\left\lVert #1\right\rVert}
\newcommand{\eps}{\varepsilon}
\newcommand{\F}{\mathrm{F}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\inner}[2]{\langle #1,\,#2\rangle}
\newcommand{\gap}{g}
\newcommand{\HH}{\mathcal H}
\newcommand{\sinT}{\sin\Theta}
\newcommand{\tp}{^{\mathsf T}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\ran}{ran}
\DeclareMathOperator{\spann}{span}

\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\title{Davis--Kahan $\sin\Theta$: what the Lean formalization proves,\\
and how it relates to Davis (1963/1970) and Yu--Wang--Samworth (2015)}

% NOTE: if you are a different model making a substantial addition and are not
% yet credited, add yourself to the author list above.
\author{Claude Opus 4.8 \and Jonathan Crall \and Claude Fable 5}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This note states, in ordinary mathematical language, the Davis--Kahan (DK)
$\sin\Theta$ theorem formalized in Lean, and locates it among the source papers.
The formalization is a ladder of Frobenius-norm $\sin\Theta$ bounds built on one
estimate, whose sharpest rung is $\norm{\sinT}_{\F}\le\norm{S-T}_{\F}/\gap$ with a
dimension-free constant.  The estimate is Davis's elementary cross-term /
Rayleigh-quotient inequality, and the separation is a hybrid gap between the
leading eigenvalues of one operator and the trailing eigenvalues of the other.
The hybrid-gap bound is the deterministic core of the Yu--Wang--Samworth (YWS)
variant; recovering their population-only gap is a further Weyl step, also
formalized.  Beyond the ladder, the formalization now also contains the exact
YWS theorem with its constant $2$ (via Hoffman--Wielandt and a von Neumann
trace inequality), the YWS aligned-basis and singular-subspace extensions,
general-interval spectral subspaces, Davis's sharper total-rotation estimate
and per-eigenvector $\sin2\theta$ and $\tan2\theta$ theorems, and the operator
Sylvester bound and the dimension-free operator-norm $\sin\Theta$ theorem it
powers, and a canonical principal-angle API.  A closing section records what is
left --- only the subspace-level $\sin2\Theta$ theorem and general unitarily
invariant norms, both deliberately deferred.
\end{abstract}

\vspace{1.5em}
% NOTE: Update permalink if underlying proof changes, so this always is up to date wrt to the referenced lean file.
\noindent Companion note to \href{https://github.com/AIQ-Kitware/aiq-dkps-formalization/blob/c17998d1279f9751610227b13bac6b3c6f263e96/ForMathlib/Analysis/InnerProductSpace/DavisKahan.lean}{ForMathlib/Analysis/InnerProductSpace/DavisKahan.lean} and its sibling files under \href{https://github.com/AIQ-Kitware/aiq-dkps-formalization/tree/c17998d1279f9751610227b13bac6b3c6f263e96/ForMathlib/Analysis/InnerProductSpace}{ForMathlib/Analysis/InnerProductSpace/}, part of a larger formalization effort at \href{https://github.com/AIQ-Kitware/aiq-dkps-formalization}{github.com/AIQ-Kitware/aiq-dkps-formalization}.



\section{Prerequisites}

This note assumes basic familiarity with the following finite-dimensional linear
algebra concepts.

\begin{itemize}
\item \textbf{Inner product spaces.}
We work over $\mathbb R$ or $\mathbb C$.  Inner products let us define
orthogonality, norms, orthonormal bases, and orthogonal projections.

\item \textbf{Self-adjoint operators.}
An operator $T$ is self-adjoint if $T=T^*$.  For complex matrices in an
orthonormal basis, this means $T$ equals its conjugate transpose; over
$\mathbb R$, this is the usual notion of a symmetric matrix.  Self-adjoint
operators have real eigenvalues and orthonormal eigenbases.

\item \textbf{Eigenvectors, eigenspaces, and spectral subspaces.}
An eigenspace is the span of all eigenvectors associated with a given
eigenvalue.  A spectral subspace is the span of eigenvectors associated with a
chosen group of eigenvalues, such as the largest $d$ eigenvalues.

\item \textbf{Orthogonal projections.}
Every subspace $U$ of a finite-dimensional inner product space has an
orthogonal projection $P_U$.  Comparing subspaces can often be reduced to
comparing their projection operators.

\item \textbf{Spectral gaps.}
A spectral gap is a positive separation between a group of eigenvalues of
interest and the rest of the spectrum.  For ordered eigenvalues
$\lambda_1\geq\lambda_2\geq\cdots\geq\lambda_n$, the leading-$d$ gap is often
$\lambda_d-\lambda_{d+1}$.

\item \textbf{Perturbation size.}
If $S$ is a perturbed version of $T$, the perturbation is $E=S-T$.  Its size is
measured using a norm, such as the operator norm or the
Frobenius/Hilbert--Schmidt norm.

\item \textbf{Frobenius/Hilbert--Schmidt norm.}
For matrices, the Frobenius norm is the square root of the sum of squared
entry magnitudes.  For operators on finite-dimensional inner product spaces,
the analogous basis-independent norm is the Hilbert--Schmidt norm.
\end{itemize}


\section{Introduction to Davis--Kahan theory}

Davis--Kahan theory is about the stability of eigenspaces under perturbation.
Suppose $T$ is a self-adjoint operator and $S=T+E$ is a nearby self-adjoint
operator. The eigenvalues and eigenvectors of $S$ may differ from those of $T$,
but if the perturbation $E=S-T$ is small, we would like to know whether the
important eigenspaces of $S$ remain close to the corresponding eigenspaces of
$T$.

This question appears naturally in principal component analysis. For example,
$T$ might be an ideal or population covariance operator, while $S$ is an
empirical covariance operator estimated from finite data. The leading
eigenvectors of $T$ define the true principal subspace, while the leading
eigenvectors of $S$ define the empirical principal subspace. Davis--Kahan theory
gives deterministic bounds on how far the empirical subspace can be from the true
subspace.

The main qualitative message is simple: eigenspaces are stable when the
perturbation is small compared to the relevant spectral gap. More explicitly,
the amount the eigenspace moves is bounded, up to constants, by the size of the
perturbation divided by the spectral gap. Thus a small perturbation causes only
a small rotation when the eigenvalues defining the subspace are well separated
from the rest of the spectrum. If the gap is tiny, however, even a small
perturbation can cause a large change, because nearby eigenspaces can mix or
exchange roles.

It is important that Davis--Kahan theory compares subspaces, not individual
eigenvectors. Individual eigenvectors are not always stable or canonical. If an
eigenvalue has multiplicity greater than one, then its eigenspace has many
possible orthonormal bases, and no single basis is preferred. Even for a
one-dimensional eigenspace, an eigenvector is only determined up to sign over
$\mathbb R$, or up to phase over $\mathbb C$. The eigenspace itself is the
stable geometric object.

There are several equivalent ways to measure the distance between two subspaces.
One way is through principal angles. If $U$ and $V$ are two $d$-dimensional
subspaces, their principal angles $\theta_1,\dots,\theta_d$ measure how far the
directions in one subspace are from the directions in the other. The notation
$\sin\Theta$ refers to the collection of numbers
$\sin\theta_1,\dots,\sin\theta_d$. A norm such as
$|\sin\Theta|_{\mathrm F}$ then measures the total misalignment between the two
subspaces.

Another way to measure subspace distance is through orthogonal projections. Let
$P_U$ and $P_V$ be the orthogonal projection operators onto $U$ and $V$. If the
subspaces are close, then the projections $P_U$ and $P_V$ are close as
operators. Many versions of Davis--Kahan theory can therefore be stated either
in terms of principal angles or in terms of projection differences such as
$|P_U-P_V|_{\mathrm F}$.

A third way is through basis overlaps. Suppose $U$ is spanned by orthonormal
vectors $u_1,\dots,u_d$, and $V$ is spanned by orthonormal vectors
$v_1,\dots,v_d$. To measure how far $U$ is from $V$, one can measure how much of
the $u_i$ directions lie outside $V$. If
$v_{d+1},\dots,v_n$ span the orthogonal complement of $V$, then the quantity
\begin{equation}
\sum_{i=1}^d \sum_{j=d+1}^n |\langle u_i, v_j\rangle|^2
\end{equation}
measures how much the $U$-subspace leaks into the complement of the
$V$-subspace. This is another expression of the same $\sin\Theta$ geometry.
Here ``trailing'' means the complementary block $v_{d+1},\dots,v_n$, not another
block of size $d$.

The current Lean formalization is closest to this basis-overlap viewpoint, with
additional lemmas connecting it to projection distance. It works in a
finite-dimensional inner product space over $\mathbb R$ or $\mathbb C$, for
self-adjoint operators $T$ and $S$. The formalized theorem fixes a block size
$d$ and compares the span of the first $d$ eigenvectors of one operator with the
span of the first $d$ eigenvectors of the other. The theorem itself does not
require the eigenvalues to be globally sorted; it only requires an explicit gap
hypothesis separating the chosen block from its complement. In the usual
application, however, the chosen block is the leading eigenspace associated with
the largest $d$ eigenvalues.

The main formalized estimate is a Frobenius/Hilbert--Schmidt version of the
Davis--Kahan $\sin\Theta$ bound. It proves that a spectral gap controls the
cross-overlap between the chosen eigenspace of one operator and the complementary
eigenspace of the other. The proof then bounds this cross-overlap by
successively larger and easier-to-use quantities: first by an off-diagonal block
of the perturbation, then by a one-sided residual, and finally by the full
Hilbert--Schmidt norm of $S-T$. Additional lemmas use Weyl-type eigenvalue
perturbation bounds to turn more standard population-gap assumptions into the
hybrid gap assumption used by the main theorem.

This means the current formalization contains a substantial Frobenius-norm core
of Davis--Kahan theory, and --- since the first version of this note --- most of
the surrounding theory as well: the ladder now runs over arbitrary index blocks
and spectral intervals, the exact Yu--Wang--Samworth theorem (population gap,
constant $2$, both the Frobenius and $\sqrt d\,$-operator-norm branches) is
formalized via Hoffman--Wielandt, the aligned-basis (Procrustes) conclusion and
the singular-subspace extension are formalized, Davis's sharper 1963
total-rotation theory is formalized together with its polar-decomposition
substrate, and Davis's per-eigenvector $\sin2\theta$ and $\tan2\theta$ theorems
are formalized in a phase-free product form, and a canonical principal-angle
API is packaged, and the dimension-free operator-norm $\sin\Theta$ theorem is
formalized on top of the operator Sylvester bound.  What is still missing is
only the subspace-level $\sin2\Theta$ theorem and general unitarily invariant
norm versions, both deliberately deferred. The purpose of this note is to
explain
exactly which part has been formalized, how it relates to the paper statements,
and what still remains to be built --- the closing section ranks the remaining
steps by difficulty and records a checked proof route for each.




\section{Setup and notation}

Let $\mathcal H$ be a finite-dimensional inner product space over
$\mathbb K\in{\mathbb R,\mathbb C}$ with $\dim\mathcal H=n$, and let
$T,S:\mathcal H\to\mathcal H$ be self-adjoint.  Write the eigenvalues and an
orthonormal eigenbasis of $T$ as $\lambda_1(T),\dots,\lambda_n(T)$ and
$u_1,\dots,u_n$, and those of $S$ as $\lambda_1(S),\dots,\lambda_n(S)$ and
$v_1,\dots,v_n$.  Fix a block size $d$.  In the intended top-eigenspace
application, the indices $1,\dots,d$ denote the leading eigenvectors, while
$d+1,\dots,n$ denote the complementary trailing eigenspace.  The formalized
hybrid-gap statement itself, however, only uses the block split
${1,\dots,d}$ versus ${d+1,\dots,n}$ together with an explicit separation
hypothesis between the two blocks; it does not require the eigenvalues to be
globally sorted unless one wants to derive that separation hypothesis from a
standard leading-eigenvalue gap.


Let $P$, $\widehat P$, and $Q=I-\widehat P$ be the orthogonal projections onto,
respectively,
\[
  \spann(u_1,\dots,u_d),\qquad
  \spann(v_1,\dots,v_d),\qquad
  \spann(v_{d+1},\dots,v_n).
\]
The first two are the leading spectral subspaces of $T$ and $S$, and $\ran Q$ is
the trailing $S$-subspace.  Let $\Theta$ be the vector of principal angles between
$\ran P$ and $\ran\widehat P$.  Two norms appear: the operator norm
$\norm{\cdot}_{\op}$ and the Frobenius, or Hilbert--Schmidt, norm
$\norm{B}_{\F}^2=\tr(B^*B)=\sum_k\norm{Be_k}^2$ over any orthonormal basis $e_k$.

The separation hypothesis is a \emph{hybrid gap}: a number $\gap>0$ with
\begin{equation}\label{eq:gap}
  \gap\le\bigl|\lambda_i(T)-\lambda_j(S)\bigr|
  \qquad\text{for all }i\le d<j.
\end{equation}
It pairs the leading eigenvalues of $T$ with the trailing eigenvalues of $S$, the
separation that Davis's estimate consumes.

\section{The formalized theorem}

The overlap $\sum_{i\le d}\sum_{j>d}|\inner{u_i}{v_j}|^2$ equals
$\norm{\sinT}_{\F}^2$, and the bounds on it form one chain of inequalities.

\begin{theorem}[Frobenius $\sin\Theta$ ladder]\label{thm:ladder}
Let $T,S$ be self-adjoint and let $\gap>0$ satisfy \eqref{eq:gap}.  Then
\begin{equation}\label{eq:ladder}
  \gap^2\,\norm{\sinT}_{\F}^2
  \;\le\;\norm{P(S-T)Q}_{\F}^2
  \;\le\;\norm{(S-T)P}_{\F}^2
  \;\le\;\norm{S-T}_{\F}^2,
\end{equation}
and consequently
\[
  \norm{\sinT}_{\F}\le\frac{\norm{S-T}_{\F}}{\gap},
  \qquad
  \norm{\widehat P-P}_{\F}\le\frac{\sqrt2\,\norm{S-T}_{\F}}{\gap},
\]
with constants independent of $n$.
\end{theorem}

The three right-hand sides of \eqref{eq:ladder} are the leading$\times$trailing
block of the perturbation, its restriction to the leading subspace (the one-sided
residual of the original DK theorem), and its full Frobenius norm; dividing by
$\gap^2$ gives three overlap bounds, of which the last is sharp.  The projector
distance and the overlap differ by an exact factor,
$\norm{\widehat P-P}_{\F}^2=2\norm{\sinT}_{\F}^2$, because
$\widehat P-P=\widehat P(I-P)-(I-\widehat P)P$ splits into two pieces with
orthogonal ranges of equal Frobenius norm, each equal to the overlap.

\begin{corollary}[operator-norm form]\label{cor:opnorm}
If $\norm{S-T}_{\op}\le\eps$, the residual has $d$ nonzero columns of norm at most
$\eps$, so $\norm{(S-T)P}_{\F}^2\le d\,\eps^2$ and
$\norm{\sinT}_{\F}^2\le d\,\eps^2/\gap^2$.  Extending to all $n$ columns gives the
cruder $\norm{\sinT}_{\F}^2\le n\,\eps^2/\gap^2$ and
$\norm{\widehat P-P}_{\F}^2\le 2n\,\eps^2/\gap^2$.  The example $S-T=\eps I$ shows
the dimension factor is unavoidable once only $\norm{S-T}_{\op}$ is controlled.
\end{corollary}

\begin{corollary}[rank and eigenvalue floor: the statistical setup]\label{cor:floor}
If $T$ is positive semidefinite of rank $d$ with nonzero eigenvalues at least
$\alpha$, and $\norm{S-T}_{\op}\le\eps\le\alpha/2$, then Weyl's inequality places
every trailing eigenvalue of $S$ below $\alpha/2$, so \eqref{eq:gap} holds with
$\gap=\alpha/2$ and $\norm{\sinT}_{\F}^2\le 4n\,\eps^2/\alpha^2$.
\end{corollary}

\section{The proof mechanism}

For eigenvectors $u_i$ of $T$ and $v_j$ of $S$, self-adjointness gives the
\emph{cross-term identity}
\begin{equation}\label{eq:cross}
  \inner{u_i}{(S-T)v_j}=\bigl(\lambda_j(S)-\lambda_i(T)\bigr)\inner{u_i}{v_j}.
\end{equation}
On a cross pair ($i\le d<j$), \eqref{eq:gap} turns this into
$\gap^2\,|\inner{u_i}{v_j}|^2\le|\inner{u_i}{(S-T)v_j}|^2$.  Summing over the
cross block gives the first inequality of \eqref{eq:ladder}, with right-hand side
$\norm{P(S-T)Q}_{\F}^2$; enlarging the column sum and then the row sum, by
Parseval in the bases $\{v_j\}$ and $\{u_i\}$, produces the remaining two rungs
and the full $\norm{S-T}_{\F}^2$.  The argument is elementary throughout, with no
resolvents or contour integrals.

\section{Relation to Davis (1963/1970)}

The estimate is Davis's~\cite{Davis1963}.  For a simple spectrum his
total-rotation bound reads, with $A\rightsquigarrow T$, $A+H\rightsquigarrow S$,
and $\mathcal C^{\perp}U$ the off-diagonal part of the matching unitary,
\[
  (\gamma')^2\,\norm{\mathcal C^{\perp}U}_{\F}^2
  \le\norm{H}_{\F}^2-\sum_i\bigl(\lambda_i-\lambda_i'\bigr)^2,
  \qquad
  \gamma_i=\min_{j\ne i}|\lambda_i-\lambda_j'|,
\]
proved by comparing $\inner{(A+H-\lambda_i')^2x_i}{x_i}$ from two sides, which is
\eqref{eq:cross}.  The formalization runs this argument on the two-block split.
The ladder \eqref{eq:ladder} realizes Davis's off-diagonal principle: the
cross-block rung $\norm{P(S-T)Q}_{\F}^2$ and the residual rung
$\norm{(S-T)P}_{\F}^2$ control the rotation by the block-mixing part of $H$,
and the sharp rung relaxes this to $\norm{S-T}_{\F}^2$.  The hybrid gap
\eqref{eq:gap} is Davis's $\gamma_i$ restricted to the cross block.  The companion
part~III of Davis and Kahan~\cite{DavisKahan1970} states the $\sin\Theta$ theorem
for every unitarily invariant norm; the Frobenius case is the instance formalized
here.

\section{Relation to Yu--Wang--Samworth (2015)}

Yu, Wang and Samworth~\cite{YWS2015} work with symmetric $\Sigma,\widehat\Sigma$
and a population eigengap $\Delta=\min(\lambda_{r-1}-\lambda_r,\
\lambda_s-\lambda_{s+1})$ formed from $\Sigma$ alone, and prove
\[
  \norm{\sinT(\widehat V,V)}_{\F}
  \le\frac{2\min\{d^{1/2}\norm{\widehat\Sigma-\Sigma}_{\op},
                   \norm{\widehat\Sigma-\Sigma}_{\F}\}}{\Delta}.
\]
Their proof is a residual sandwich
$\Delta\norm{\sinT}_{\F}\le\norm{R}_{\F}\le
2\min\{d^{1/2}\norm{\cdot}_{\op},\norm{\cdot}_{\F}\}$ with
$R=\widehat V\Lambda-\Sigma\widehat V$: a deterministic separation estimate on the
left, a perturbation estimate (Weyl and Hoffman--Wielandt) on the right, and the
factor $2$ paid for moving to the population-only gap $\Delta$.

The formalization keeps these two pieces separate.  The hybrid-gap bound, the
sharp rung of \eqref{eq:ladder}, is the deterministic separation estimate with
constant $1$: the hybrid gap already places the trailing eigenvalues, so the
Hoffman--Wielandt displacement term drops out.  The lemma
\texttt{gap\_of\_eigengap} supplies the Weyl step, converting a spectral gap of
$T$ (leading eigenvalues $\ge a$, trailing $\le b$) into the hybrid gap
$(a-b)-\eps$; YWS's $\Delta=\lambda_d-\lambda_{d+1}$ is the case $a-b$, and
\texttt{gap\_of\_rank\_floor} the case $a=\alpha$, $b=0$.  Their $d^{1/2}$
operator-norm branch is Corollary~\ref{cor:opnorm},
$\norm{\sinT}_{\F}\le d^{1/2}\eps/\gap$.

Since the first version of this note, the \emph{exact} YWS theorem --- the
population-only gap $\Delta$ with the constant $2$, in both the Frobenius and
the $d^{1/2}$-operator-norm branches --- has also been formalized, following
YWS's own residual sandwich through Hoffman--Wielandt rather than the Weyl
bridge; see \S\ref{sec:extensions}.

\section{Extensions formalized since the first version}\label{sec:extensions}

The first version of this note stopped at the ladder of
Theorem~\ref{thm:ladder}.  Most of the gaps it recorded have since been closed;
this section states each result in ordinary language.  Lean names for
everything here appear in the second dictionary table of
\S\ref{sec:dictionary}.

\paragraph{General blocks and spectral intervals.}
The engine behind Theorem~\ref{thm:ladder} never uses that the selected block
is the \emph{leading} block.  It is now stated for two independent index sets
$s$ (rows, for $T$) and $t$ (columns, for $S$) under the cross-gap hypothesis
$\gap\le|\lambda_i(T)-\lambda_j(S)|$ for $i\in s$, $j\notin t$; the sorted
$d$-block statements are one-line corollaries.  On top of this sit
interval-selected subspaces: if the $T$-block is
$\{i:\lambda_i(T)\in[a,b]\}$ and every trailing eigenvalue of $S$ avoids the
enlarged interval $(a-\gap,\,b+\gap)$, the sharp rung applies verbatim; a
two-sided Weyl bridge converts a population interval gap $\delta$ of $T$ plus
$\norm{S-T}_{\op}\le\eps<\delta$ into the hypothesis with $\gap=\delta-\eps$.
This closes the ``general-interval subspaces'' item of the first version.

\paragraph{Hoffman--Wielandt and the exact YWS theorem.}
The Weyl bridge alone cannot reach YWS's constant-$2$ Frobenius branch when
$\norm{E}_{\op}>\Delta/2$ (the hybrid gap $\Delta-\norm{E}_{\op}$ degenerates).
The formalization now follows YWS's own residual sandwich.  The
Hoffman--Wielandt inequality
\[
  \sum_i\bigl(\lambda_i(T)-\lambda_i(S)\bigr)^2\;\le\;\norm{S-T}_{\F}^2
\]
is proved via a von Neumann-type trace inequality
$\sum_k\lambda_k(T)\,\mathrm{re}\inner{u_k}{Su_k}\le\sum_i\lambda_i(T)\lambda_i(S)$,
itself obtained from the Schur--Horn doubly stochastic matrix of the two
eigenbases, Birkhoff's theorem, and the rearrangement inequality.  With
Hoffman--Wielandt in hand, the residual columns
$R_j=\lambda_j(T)\,\widehat v_j-T\widehat v_j$ satisfy the two-sided estimate
$\Delta^2\norm{\sinT}_{\F}^2\le\sum_j\norm{R_j}^2\le
\bigl(2\min\{d^{1/2}\norm{E}_{\op},\norm{E}_{\F}\}\bigr)^2$, giving the exact
YWS headline
\[
  \norm{\sinT}_{\F}\;\le\;\frac{2\min\{d^{1/2}\norm{E}_{\op},\ \norm{E}_{\F}\}}{\Delta}
\]
with the population-only gap $\Delta$ and YWS's constant $2$, in both branches.

\paragraph{The aligned-basis (Procrustes) bound.}
There is an orthonormal family $w_1,\dots,w_d$ spanning the sample block
$\spann(\widehat v_1,\dots,\widehat v_d)$ with
\[
  \sum_{j\le d}\norm{w_j-u_j}^2\;\le\;2\,\norm{\sinT}_{\F}^2,
  \qquad\text{hence}\qquad
  \Bigl(\sum_j\norm{w_j-u_j}^2\Bigr)^{1/2}
  \le\frac{2^{3/2}\min\{d^{1/2}\norm{E}_{\op},\norm{E}_{\F}\}}{\Delta},
\]
which is YWS's $\norm{\widehat VO-V}_{\F}$ conclusion in operator language
($w=\widehat VO$).  The rotation $O$ is the \emph{kernel-completed unitary}
polar factor of the $d\times d$ overlap operator $M_{ij}=\inner{u_i}{\widehat
v_j}$ --- genuinely unitary even when $M$ is singular (a principal angle of
$\pi/2$), which is what lets the proof avoid the singular value decomposition
entirely: the Procrustes cross term becomes the exact identity
$\mathrm{re}\tr(O^{*}M)=\tr\lvert M\rvert=\sum_i\sigma_i(M)$, and the
elementary chain $\sum_i\sigma_i\ge\sum_i\sigma_i^2=d-\norm{\sinT}_{\F}^2$
(each $\sigma_i\le 1$ because $M$ is a contraction) finishes exactly as in
YWS's identity $2d-2\sum\cos\theta_i\le2\sum\sin^2\theta_i$.

\paragraph{The singular-subspace extension.}
For rectangular $A,\widehat A$ the right-singular subspaces are the spectral
subspaces of the Gram operators $A^{*}A,\widehat A^{*}\widehat A$, so the
symmetric theory applies once the Gram perturbation is controlled:
$\norm{(\widehat A^{*}\widehat A-A^{*}A)x}\le(\norm{A}+\norm{\widehat
A})\,\norm{\widehat A-A}\,\norm{x}$, Weyl for squared singular values
$|\sigma_k(\widehat A)^2-\sigma_k(A)^2|\le(\norm{A}+\norm{\widehat A})\eps$,
and the YWS engine on the Gram pair give
$\Gamma^2\,\norm{\sinT}_{\F}^2\le 4d\bigl((\norm{A}+\norm{\widehat
A})\eps\bigr)^2$ for the population singular-value-squared gap $\Gamma$.  The
left-singular mirror follows by swapping $A\leftrightarrow A^{*}$.

\paragraph{Davis's sharper total-rotation estimate.}
Davis's Theorem~3.2, the refinement
\[
  (\gamma')^2\sum_i\sin^2\theta_i+\sum_i(\lambda_i-\lambda_i')^2
  \le\norm{H}_{\F}^2,
  \qquad
  (\gamma')^2=\min_i\bigl\{\gamma_i^2-(\lambda_i-\lambda_i')^2\bigr\},
\]
which subtracts the eigenvalue displacement, and its corollary
$(\gamma')^2\norm{\mathcal C^{\perp}U}_{\F}^2\le 2\norm{\mathcal
C^{\perp}H}_{\F}^2$ combining Davis's Schur--Horn eigenvalue-change lower bound
(his Theorem~4.1).  The supporting machinery --- the positive square root, a
partial-isometry API, the operator polar decomposition $A=U\lvert A\rvert$
(with a $\mathbb C$/\texttt{CFC.abs} headline), the kernel-completed
\emph{unitary} polar decomposition used above, and Davis's \S2 canonical
intertwining unitary of two complete orthogonal projection families --- lives
in \texttt{PositiveSqrt.lean}, \texttt{PartialIsometry.lean},
\texttt{PolarDecomposition.lean}, \texttt{IntertwiningUnitary.lean}; the
rotation theorems themselves are in \texttt{RotationBound.lean}.

\paragraph{The per-eigenvector $\sin2\theta$ theorem.}
Davis's sharp two-subspace estimate, in product form: if the quadratic form of
the symmetric $T$ is $\ge b\norm{\cdot}^2$ on a $T$-invariant subspace $U$ and
$\le a\norm{\cdot}^2$ on $U^{\perp}$, $x$ is a unit eigenvector of $T+H$ with
\emph{no assumption on the location of its eigenvalue}, $\norm{H}_{\op}\le\eps$,
and $P=P_U$, then
\[
  (b-a)\,\norm{Px}\,\norm{x-Px}\;\le\;\eps,
  \qquad\text{i.e.}\qquad
  \sin2\theta\;\le\;\frac{2\eps}{b-a}.
\]
The formalized proof is phase-free and radical-free: pairing the eigenvector
equation with $y=Px$ and $z=x-Px$ and taking real parts eliminates both the
eigenvalue and the complex phase
($\mathrm{re}(c^2w-s^2\bar w)=(c^2-s^2)\,\mathrm{re}\,w$ identically, with
$c=\norm{y}$, $s=\norm{z}$), and the classical half-angle rotation is realized
by the two test vectors $s(c-s)\,y+c(c+s)\,z$ and $-s(c+s)\,y+c(c-s)\,z$ with
polynomial coefficients, using $1\mp2cs=(c\mp s)^2$.  The primary Lean
statement is in orthogonal-decomposition form and assumes no orthogonal
projection, completeness, or finite dimension; the projection and
$\arccos$-angle forms are corollaries.

\paragraph{The per-eigenvector $\tan2\theta$ theorem.}
Under the same setup but with the \emph{vanishing-pinch} hypotheses --- $H$ has
no diagonal blocks with respect to $U\oplus U^{\perp}$, i.e.\
$\inner{u}{Hu'}=0$ for $u,u'\in U$ and $\inner{w}{Hw'}=0$ for $w,w'\in
U^{\perp}$ --- the same key identity collapses, with the two diagonal
$H$-terms gone, to
$\norm{z}^2\mathrm{re}\inner{y}{Ty}-\norm{y}^2\mathrm{re}\inner{z}{Tz}
=(\norm{y}^2-\norm{z}^2)\,\mathrm{re}\inner{y}{Hz}$, and bounding the single
mixed term directly (no rotation, no half-angle) gives
\[
  (b-a)\,\norm{Px}\,\norm{x-Px}\;\le\;\bigl|\norm{Px}^2-\norm{x-Px}^2\bigr|\,\eps,
  \qquad\text{i.e.}\qquad
  \tan2\theta\;\le\;\frac{2\eps}{b-a},
\]
with \emph{no smallness assumption} on the perturbation (the diagonal-block
hypothesis replaces it).  The product form carries no $\theta\ne\pi/4$ side
condition.

\paragraph{The Sylvester (Lyapunov) solution bound.}
The analytic engine for the dimension-free operator-norm $\sin\Theta$ theorem
is in place: if $A,B$ are bounded symmetric with quadratic forms
$\delta$-coercive ($\delta\norm{x}^2\le\mathrm{re}\inner{Ax}{x}$ and likewise
for $B$) and $AX+XB=Y$, then $\norm{X}\le\norm{Y}/(2\delta)$; equivalently, in
separated form, if the form of $A$ sits above $c+\gap$ and that of $B$ below
$c$ and $AX-XB=Y$, then $\norm{X}\le\norm{Y}/\gap$.  The proof is a purely
algebraic one-inequality argument (no resolvents, contour integrals, or
Bochner integrals), and holds with no finite-dimensionality or completeness
assumptions.  It is the engine of the operator-norm $\sin\Theta$ theorem
recorded next.

\paragraph{The canonical principal-angle API.}
For two orthonormal families $u,v:\mathrm{Fin}\,d\to\HH$ spanning subspaces
$U,V$, the cosines of the principal angles are defined as the singular values
of the flat overlap operator (matrix $\inner{u_i}{v_j}$), and $\sin\Theta$'s
squared Frobenius norm as $\sum_i(1-\cos^2\theta_i)$.  The formalization records
that the cosines lie in $[0,1]$ and are sorted, are \emph{symmetric} in the two
families (the swapped overlap operator is the adjoint, and adjoints share
singular values --- this is exactly where $\sigma(M^{*})=\sigma(M)$ is used, and
the reason it was proved), and satisfy $\norm{\sinT}_{\F}^2=d-\mathrm{overlap}$
with $\mathrm{overlap}=\sum_j\sum_i|\inner{u_i}{v_j}|^2$.  The Yu--Wang--Samworth
aligned-basis bound is restated in this vocabulary as
$\sum_j\norm{w_j-u_j}^2\le2\norm{\sinT}_{\F}^2$.  The singular-value symmetry
$\sigma(A^{*})=\sigma(A)$ (square case) is itself a new addition, proved through
the invariance of a symmetric operator's sorted eigenvalues under unitary
conjugation ($AA^{*}=U(A^{*}A)U^{-1}$ with $U$ the kernel-completed unitary
polar factor).

\paragraph{The dimension-free operator-norm $\sin\Theta$ theorem.}
The headline $\norm{\sinT}_{\op}\le\norm{S-T}_{\op}/\gap$ is now formalized,
in the vocabulary $\norm{\widehat Q\,P}_{\op}\le\eps/\gap$ where $P$ projects
onto a $T$-invariant subspace $U$ whose quadratic form sits above $c+\gap$ and
$\widehat Q$ onto an $S$-invariant subspace $V$ whose form sits below $c$ (that
operator norm \emph{is} $\norm{\sinT}_{\op}$).  The proof compresses nothing:
on the full space, with $X=P\widehat Q$ and the scalar-extended operators
$A=TP+(c+\gap)(I-P)$, $B=SQ+c(I-Q)$ (globally $(c+\gap)$-coercive and
$c$-bounded by the block decomposition, since $U,U^{\perp}$ are $T$-invariant
and $V,V^{\perp}$ are $S$-invariant), the projection algebra gives the exact
Sylvester relation $AX-XB=P(T-S)Q$, and the operator Sylvester bound above
finishes; $\norm{\widehat Q\,P}=\norm{P\,\widehat Q}$ by self-adjointness of
the projections.  No dimension factor appears, and (unlike the Frobenius
ladder) no eigenbasis or Courant--Fischer machinery is used --- the two
quadratic-form hypotheses are taken abstractly.

\paragraph{The certified principal-angle dictionary.}
The extreme singular values are now variationally characterized and attained
($\sigma_{\min}\norm{x}\le\norm{Ax}\le\sigma_0\norm{x}$, with unit witnesses
for both), and on top of this sits the operator-norm identification
\[
  \norm{P_{W^{\perp}}\,P_U}_{\op}\;=\;\sqrt{1-\cos^2\theta_{\max}}
  \;=\;\sin\theta_{\max},
\]
certifying that the dimension-free operator-norm theorem above literally
bounds the largest principal-angle sine.  Every abstract invariant-subspace
theorem (operator-norm $\sin\Theta$, per-vector $\sin2\theta$/$\tan2\theta$)
also received its literature-facing corollary with sorted-eigenvalue
hypotheses, and the Frobenius encodings in this repository --- the cross-block
overlap sum, $\norm{\sinT}_{\F}^2=d-\norm{\cos\Theta}_{\F}^2$, and the
projector distance $\norm{\widehat P-P}_{\F}^2$ --- are proved pairwise equal
on eigenbasis blocks.

\paragraph{Ky Fan sums, unitarily invariant norms, and Fan dominance.}
The core of the deferred ``unitarily invariant norm'' project is now built.
The singular-value calculus was completed (unitary invariance on both sides,
nonnegative real scaling, and the bounded-factor domination
$\sigma_i(CA)\le c\,\sigma_i(A)$, via a new Loewner-monotonicity lemma for
sorted eigenvalues and the sorted-eigenvalue uniqueness principle).  On it
stand the Ky Fan trace inequality
$\sum_i\mathrm{re}\inner{Sw_i}{w_i}\le\sum_{i<k}\lambda_i(S)$, the Ky Fan
variational principle
$\sum_{i<k}\sigma_i(A)=\max\,\mathrm{re}\sum_i\inner{u_i}{Av_i}$ (both
directions), and weak majorization
$\sigma(A+B)\prec_w\sigma(A)+\sigma(B)$ --- the triangle inequality for every
Ky Fan norm at once.  A \texttt{UnitarilyInvariantNorm} structure (seminorm
axioms plus two-sided unitary invariance) then receives the full symmetric
gauge apparatus: an operator SVD $A=U\,\mathrm{diag}(\sigma(A))\,V$ relative
to \emph{any} fixed orthonormal basis, the gauge representation
$N(A)=\Phi_N(\sigma(A))$, and the \textbf{Fan dominance principle} --- Ky Fan
domination for all $k$ implies $N(A)\le N(B)$ for every unitarily invariant
norm.  The classical route (weak-majorization completion, then
Hardy--Littlewood--P\'olya via Birkhoff) is bypassed: a T-transform descent is
run \emph{directly on the gauge}, each step averaging $y$ with a transposition
of itself at the cost of one triangle inequality, one homogeneity, and one
swap invariance.  Star invariance $N(A^{*})=N(A)$ comes free from the gauge
representation.

\paragraph{The part-III $\sin\Theta$ theorem, every unitarily invariant norm.}
With the Fan-dominance library in hand, the Davis--Kahan~(1970) part-III
$\sin\Theta$ theorem is now formalized at full generality:
\[
  N\bigl(\widehat Q\,P\bigr)\;\le\;N(S-T)\,/\,g
  \qquad\text{for every unitarily invariant norm } N,
\]
under exactly the operator-norm theorem's hypotheses ($P,\widehat Q$ the
projections onto the separated invariant subspaces).  The engine is an
\emph{abstract} Sylvester bound: the same integral-free absorption identity
$(\norm{A}+\norm{B})X = Y + (\norm{A}\cdot1-A)X + X(\norm{B}\cdot1-B)$, but
now applied under any operator seminorm $N$ carrying the ideal property
$N(CX)\le\norm{C}\,N(X)$ --- and every unitarily invariant norm does, by Fan
dominance applied to $\sigma_i(CX)\le\norm{C}\,\sigma_i(X)$.  Because the
identity is an equation of \emph{operators}, $N$ applies to it directly, with
no pointwise estimate.  The norm-free construction of $A,B,X,Y$ is shared
verbatim with the operator-norm proof; only the final estimate differs.  The
operator norm is exhibited as an instance (\texttt{opNorm}), so the theorem is
not vacuous, and the Frobenius and operator-norm ladders are its instances.

\paragraph{The subspace $\sin2\Theta$ theorem, every unitarily invariant norm.}
The first of the three remaining part-III theorems is now formalized: if the
form of $T$ sits above $b$ on a $T$-invariant subspace $U$ and below $a$ on
$U^{\perp}$ (the gap constrains $T$ \emph{alone}), and $V$ is any
$S$-invariant subspace, then for every unitarily invariant norm
\[
  N\bigl(Q\,\widehat P\,P\bigr)\;\le\;N(S-T)\,/\,(b-a),
  \qquad P=P_U,\; Q=1-P,\; \widehat P=P_V .
\]
Since $2\,Q\widehat P P$ has singular values $\sin2\theta_i$, this is
$\norm{\sin2\Theta}\le 2\norm{S-T}/(b-a)$ --- with \emph{no smallness
assumption} on the perturbation and \emph{no spectral-location constraint} on
$V$, strictly more general than the classical statement.  The proof is the
mirror reduction of Davis--Kahan's own \S8: reflect $T$ through the perturbed
subspace ($J=2\widehat P-1$, $T'=JTJ$) and apply the $\sin\Theta$ theorem
above to the pair $(T,T')$, which is separated by $T$'s own gap; the resulting
cross-projection is $J$-conjugate to $Q J P=2\,Q\widehat P P$, and
$N(T'-T)\le 2N(S-T)$ because $J$ commutes with $S$.  A sharper
\emph{mirror-defect} form is recorded separately:
$2N(Q\,P_W\,P)\le N(JTJ-T)/(b-a)$ for an \emph{arbitrary} mirror subspace
$W$ --- no second operator at all.

\section{What remains, ranked by difficulty}\label{sec:remains}

Each downstream consumer in this repository is already served by the
Frobenius and operator-norm results above.  A full roadmap for closing them --- the Davis--Kahan~III quartet
($\sin\Theta$, $\sin2\Theta$, $\tan\Theta$, $\tan2\Theta$) at the subspace
level in every unitarily invariant norm, via a Ky Fan / Fan-dominance
mini-library, together with the certified principal-angle dictionary --- is
maintained, with per-step proof routes and difficulty grades, in
\texttt{dev/davis-kahan-expert-completion-plan.md}; the executed plan for
everything above is \texttt{dev/davis-kahan-gap-closure-plan.md} (v6, closed).

\begin{enumerate}
\item \emph{Subspace-level $\tan2\Theta$ (vanishing pinch) and $\tan\Theta$
  (difficulty 4.5--5/5 --- the last two part-III theorems).}  The subspace
  $\sin2\Theta$ theorem is now formalized (see \S\ref{sec:extensions});
  the mirror reduction that proves it yields \emph{sine-type absolute}
  bounds, while $\tan2\Theta$ under the vanishing-pinch hypothesis is a
  \emph{relative} bound (angles past $\pi/4$ are allowed), so it needs
  different technology --- likely the per-vector key identity summed with the
  diagonal blocks hypothesized away.  $\tan\Theta$ (one exact spectral
  subspace against an arbitrary test subspace, residual on the right-hand
  side) carries the highest statement risk; per-step routes and descopes are
  in the expert-completion plan, steps G2--G3.

\end{enumerate}

\section{Dictionary to the Lean source}\label{sec:dictionary}

The core ladder declarations live in
\texttt{ForMathlib/Analysis/InnerProductSpace/DavisKahan.lean}
(namespace \texttt{ForMathlib}).

\begin{center}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}p{0.42\linewidth}l@{}}
\toprule
Statement here & Lean name \\
\midrule
Cross-term identity \eqref{eq:cross}
  & \texttt{inner\_eigenvectorBasis\_map\_sub\_eigenvectorBasis} \\
Parseval identities (full sum / single row)
  & \texttt{sum\_sq\_norm\_inner\_\dots\_map\_sub\_eq}\,/\,\texttt{\dots\_eq\_row} \\
Dimension-factor step $\norm{S-T}_{\F}^2\le n\eps^2$
  & \texttt{sum\_norm\_eigenvectorBasis\_map\_sub\_sq\_le} \\
Ladder, cross-block rung $\norm{P(S-T)Q}_{\F}^2/\gap^2$
  & \texttt{sum\_cross\_\dots\_sq\_le\_offDiag} \\
Ladder, residual rung $\norm{(S-T)P}_{\F}^2/\gap^2$
  & \texttt{sum\_cross\_\dots\_sq\_le\_residual} \\
Ladder, sharp rung $\norm{S-T}_{\F}^2/\gap^2$ (Thm~\ref{thm:ladder})
  & \texttt{sum\_cross\_\dots\_sq\_le\_hilbertSchmidt} \\
Cor.~\ref{cor:opnorm}, $d^{1/2}$ branch $d\,\eps^2/\gap^2$
  & \texttt{sum\_cross\_\dots\_sq\_le\_opNorm} \\
Cor.~\ref{cor:opnorm}, crude $n\eps^2/\gap^2$
  & \texttt{sum\_cross\_\dots\_sq\_le} \\
Population-gap Weyl bridge $\Rightarrow\gap=(a-b)-\eps$
  & \texttt{gap\_of\_eigengap} \\
Projector identity $\norm{\widehat P-P}_{\F}^2=2\cdot\text{overlap}$
  & \texttt{sum\_norm\_sub\_starProjection\_span\_sq\_eq} \\
Projector form of Thm~\ref{thm:ladder}
  & \texttt{sum\_norm\_sub\_starProjection\_span\_sq\_le\_hilbertSchmidt} \\
Projector, crude $2n\eps^2/\gap^2$
  & \texttt{sum\_norm\_sub\_starProjection\_span\_sq\_le} \\
Weyl bridge, rank/floor $\Rightarrow\gap=\alpha/2$
  & \texttt{gap\_of\_rank\_floor} \\
Cor.~\ref{cor:floor}, rank/floor $4n\eps^2/\alpha^2$
  & \texttt{sum\_cross\_\dots\_sq\_le\_of\_rank\_floor} \\
\bottomrule
\end{tabular}
\end{center}

The projections $P,\widehat P$ are Mathlib's \texttt{Submodule.starProjection}
of the spans of the selected eigenvectors; the Frobenius norm on the left of the
projector bound is written in Lean as the basis sum
$\sum_k\norm{(\widehat P-P)u_k}^2=\norm{\widehat P-P}_{\F}^2$.

The extensions of \S\ref{sec:extensions} live in sibling files of the same
directory (namespace \texttt{ForMathlib} throughout):

\begin{center}
\footnotesize
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}p{0.40\linewidth}p{0.56\linewidth}@{}}
\toprule
Statement (\S\ref{sec:extensions}) & Lean name (file) \\
\midrule
General-block ladder (rows $s$, columns $t$)
  & \texttt{sum\_cross\_\dots\_sq\_le\_\{offDiag,residual,hilbertSchmidt\}\_block} (\texttt{DavisKahan.lean}) \\
Interval subspaces, sharp rung / two-sided Weyl bridge
  & \texttt{sum\_cross\_interval\_sq\_le\_hilbertSchmidt(\_of\_eigengap)},
    \texttt{notMem\_Ioo\_eigenvalues\_of\_notMem\_Ioo} (\texttt{DavisKahan.lean}) \\
Rearrangement / Birkhoff bilinear / von Neumann trace
  & \texttt{sum\_mul\_comp\_perm\_le\_sum\_mul\_of\_antitone},
    \texttt{sum\_mul\_sum\_mul\_le\_sum\_mul\_of\_antitone},
    \texttt{sum\_eigenvalues\_mul\_re\_inner\_self\_le} (\texttt{HoffmanWielandt.lean}) \\
Hoffman--Wielandt
  & \texttt{sum\_sq\_eigenvalues\_sub\_le\_sum\_sq\_norm\_apply} (\texttt{HoffmanWielandt.lean}) \\
Exact YWS, Frobenius branch ($\Delta^2\cdot\text{overlap}\le4\norm{E}_{\F}^2$)
  & \texttt{sq\_gap\_mul\_sum\_cross\_le\_of\_population\_gap},
    \texttt{sqrt\_sum\_cross\_le\_of\_population\_gap} (\texttt{YuWangSamworth.lean}) \\
Exact YWS, $d^{1/2}$ operator-norm branch
  & \texttt{sq\_gap\_mul\_sum\_cross\_le\_of\_population\_gap\_opNorm} (\texttt{YuWangSamworth.lean}) \\
Aligned-basis bound $\sum_j\norm{w_j-u_j}^2\le2\norm{\sinT}_{\F}^2$
  & \texttt{sum\_sq\_norm\_aligned\_le} (\texttt{AlignedBasis.lean}) \\
Overlap operator, contraction, $\sum\sigma^2$, $d-\text{overlap}\le\sum\sigma$
  & \texttt{overlapOp}, \texttt{overlapOp\_contraction},
    \texttt{sum\_sq\_singularValues\_overlapOp},
    \texttt{sum\_overlap\_le\_sum\_singularValues} (\texttt{AlignedBasis.lean}) \\
Singular-value glue ($\norm{\cdot}_{\F}^2=\sum\sigma^2$, $\sigma\le1$,
$\tr\lvert A\rvert=\sum\sigma$)
  & \texttt{sum\_sq\_singularValues},
    \texttt{singularValues\_le\_one\_of\_contraction},
    \texttt{sum\_re\_inner\_abs\_self\_eq\_sum\_singularValues} (\texttt{SingularSubspace.lean}) \\
Gram perturbation / singular-value Weyl / YWS Theorem 3
  & \texttt{norm\_gram\_sub\_gram\_apply\_le},
    \texttt{abs\_sq\_singularValues\_sub\_le},
    \texttt{sq\_gap\_mul\_sum\_cross\_singularVectors\_le} (\texttt{SingularSubspace.lean}) \\
Kernel-completed unitary polar decomposition
  & \texttt{polarUnitary}, \texttt{polar\_decomposition\_unitary} (\texttt{PolarDecomposition.lean}) \\
Davis Thm 3.2 + off-diagonal corollary
  & \texttt{RotationBound.lean} (see the file header for the statement map) \\
Per-eigenvector $\sin2\theta$: decomposition / projection / angle form,
$U^{\perp}$-invariance helper
  & \texttt{sin\_two\_theta\_le\_of\_mem}, \texttt{sin\_two\_theta\_le},
    \texttt{sin\_two\_arccos\_le},
    \texttt{map\_mem\_orthogonal\_of\_forall\_map\_mem} (\texttt{RotationSharp.lean}) \\
Per-eigenvector $\tan2\theta$ (vanishing pinch): decomposition / projection form
  & \texttt{tan\_two\_theta\_le\_of\_mem}, \texttt{tan\_two\_theta\_le}
    (\texttt{RotationSharp.lean}) \\
Sylvester bound, coercive / separated form
  & \texttt{opNorm\_le\_div\_of\_comp\_add\_comp\_eq},
    \texttt{opNorm\_le\_div\_of\_comp\_sub\_comp\_eq} (\texttt{SylvesterBound.lean}) \\
Operator-norm $\sin\Theta$ / projection--operator commutation
  & \texttt{norm\_starProjection\_comp\_starProjection\_le},
    \texttt{starProjection\_comp\_toContinuousLinearMap\_comm}
    (\texttt{SinThetaOpNorm.lean}) \\
$\sigma(A^{*})=\sigma(A)$ / eigenvalue conjugation-invariance
  & \texttt{singularValues\_adjoint}, \texttt{eigenvalues\_conj\_unitary}
    (\texttt{SingularSubspace.lean}) \\
Principal-angle cosines / $\sin\Theta$ / symmetry / overlap bridge
  & \texttt{cosPrincipalAngles}, \texttt{sinThetaSq},
    \texttt{cosPrincipalAngles\_comm}, \texttt{sinThetaSq\_eq\_sub\_overlap}
    (\texttt{PrincipalAngles.lean}) \\
Extreme singular values: bounds and attainment
  & \texttt{singularValues\_last\_mul\_norm\_le},
    \texttt{norm\_apply\_le\_singularValues\_zero\_mul},
    \texttt{exists\_norm\_apply\_eq\_singularValues\_last}\,/\,\texttt{\dots\_zero}
    (\texttt{SingularSubspace.lean}) \\
Identification $\norm{P_{W^{\perp}}P_U}_{\op}=\sin\theta_{\max}$
  & \texttt{norm\_orthogonal\_starProjection\_comp\_starProjection}
    (\texttt{PrincipalAngles.lean}) \\
Spectral (sorted-eigenvalue) corollaries; $\sin\theta_{\max}\le\eps/\gap$
  & \texttt{norm\_starProjection\_comp\_starProjection\_le\_of\_eigenvalues},
    \texttt{sin\_two\_theta\_le\_of\_eigenvalues},
    \texttt{tan\_two\_theta\_le\_of\_eigenvalues},
    \texttt{sqrt\_one\_sub\_sq\_cosPrincipalAngles\_le}
    (\texttt{SinThetaOpNorm.lean}) \\
Frobenius-encoding coherence (block families)
  & \texttt{sinThetaSq\_blockFamily\_eq\_sum\_cross},
    \texttt{sum\_norm\_sub\_starProjection\_sq\_eq\_two\_mul\_sinThetaSq}
    (\texttt{PrincipalAngles.lean}) \\
Loewner monotonicity / sorted-eigenvalue uniqueness
  & \texttt{eigenvalues\_le\_eigenvalues\_of\_re\_inner\_le},
    \texttt{eigenvalues\_eq\_of\_eigenbasis} (\texttt{CourantFischer.lean}) \\
Singular values under unitaries / scaling / bounded factors
  & \texttt{singularValues\_unitary\_comp}\,/\,\texttt{\_comp\_unitary},
    \texttt{singularValues\_real\_smul},
    \texttt{singularValues\_comp\_le}\,/\,\texttt{'} (\texttt{KyFan.lean}) \\
Ky Fan trace inequality / variational principle
  & \texttt{sum\_re\_inner\_le\_sum\_eigenvalues\_top},
    \texttt{re\_sum\_inner\_map\_le\_sum\_singularValues},
    \texttt{exists\_orthonormal\_re\_sum\_inner\_map\_eq} (\texttt{KyFan.lean}) \\
Ky Fan sums; weak majorization $\sigma(A+B)\prec_w\sigma(A)+\sigma(B)$
  & \texttt{kyFanSum}, \texttt{kyFanSum\_add\_le} (\texttt{KyFan.lean}) \\
Diagonal operators; operator SVD $A=U\,\mathrm{diag}(\sigma(A))\,V$
  & \texttt{diagOp}, \texttt{singularValues\_diagOp},
    \texttt{exists\_unitary\_diagOp\_factorization}
    (\texttt{UnitarilyInvariantNorm.lean}) \\
UI-norm structure; gauge representation $N=\Phi_N\circ\sigma$
  & \texttt{UnitarilyInvariantNorm}, \texttt{apply\_eq\_gauge}
    (\texttt{UnitarilyInvariantNorm.lean}) \\
T-transform descent on the gauge; \textbf{Fan dominance}; $N(A^{*})=N(A)$
  & \texttt{gauge\_le\_gauge\_of\_prefix\_sums\_le},
    \texttt{apply\_le\_of\_kyFanSum\_le}, \texttt{apply\_adjoint}
    (\texttt{UnitarilyInvariantNorm.lean}) \\
Operator-ideal property $N(CX)\le\norm{C}N(X)$ (both factors)
  & \texttt{apply\_comp\_le}, \texttt{apply\_comp\_le'}
    (\texttt{UnitarilyInvariantNorm.lean}) \\
Abstract Sylvester bound (coercive / separated), any ideal seminorm
  & \texttt{le\_div\_of\_comp\_add\_comp\_eq},
    \texttt{le\_div\_of\_comp\_sub\_comp\_eq} (\texttt{SylvesterBound.lean}) \\
Norm-free Davis--Kahan setup (shared with the op-norm proof)
  & \texttt{exists\_isSymmetric\_comp\_sub\_comp\_eq}
    (\texttt{SinThetaOpNorm.lean}) \\
Part-III $\sin\Theta$, every UI norm; operator-norm instance
  & \texttt{apply\_starProjection\_comp\_starProjection\_le},
    \texttt{opNorm} (\texttt{SinThetaUINorm.lean}) \\
Subspace $\sin2\Theta$, every UI norm; mirror-defect form
  & \texttt{sin\_two\_theta\_starProjection\_le},
    \texttt{sin\_two\_theta\_reflection\_le}
    (\texttt{SinTwoThetaUINorm.lean}) \\
\bottomrule
\end{tabular}
\end{center}

\begin{thebibliography}{9}
\bibitem{Davis1963}
Chandler Davis.
\newblock \href{https://doi.org/10.1016/0022-247X(63)90001-5}{The rotation of eigenvectors by a perturbation}.
\newblock \emph{J.\ Math.\ Anal.\ Appl.}\ \textbf{6} (1963), 159--173.

\bibitem{DavisKahan1970}
Chandler Davis and W.\ M.\ Kahan.
\newblock \href{https://doi.org/10.1137/0707001}{The rotation of eigenvectors by a perturbation.\ III}.
\newblock \emph{SIAM J.\ Numer.\ Anal.}\ \textbf{7} (1970), no.\ 1, 1--46.

\bibitem{YWS2015}
Yi Yu, Tengyao Wang, and Richard J.\ Samworth.
\newblock \href{https://doi.org/10.1093/biomet/asv008}{A useful variant of the Davis--Kahan theorem for statisticians}.
\newblock \emph{Biometrika} \textbf{102} (2015), no.\ 2, 315--323.
\newblock arXiv:\href{https://arxiv.org/abs/1405.0680}{1405.0680}.
\end{thebibliography}

\end{document}
